---
title: "Customer Segmentation Analysis"
author: "Femi Ogungbade"
date: "8/25/2023"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Intro
Customer segmentation is a marketing strategy that involves dividing a company's customer base into distinct groups or segments based on certain shared characteristics or behaviors. The goal of customer segmentation is to better understand customers' needs, preferences, and behaviors in order to tailor marketing efforts, products, and services to meet their specific requirements.

The kpi analysis and rfm segmentation has been done using sql and you can find it in the repo under sales_data_analysis.sql

The inspiration for this post was post was gotten from this [kaggle notebook](https://www.kaggle.com/code/alekhyabotta/customer-segmentation/notebook) written in python. 

### Load Library and Data. 
1. Start off loading the necessary libraries, mdsr is optional.
2. The csv is available in the repo.

```{r Load libraries and read data}

library(tidyverse)
# library(mdsr)
# load data from csv
rfm_table = read.csv("rfm_summary.csv")

# Increasing Recency values by 1 because we would log transform the data later
# this is to avoid getting any Inf values in the log transformed data
rfm_data = rfm_table %>% 
  mutate(Recency = Recency + 1) %>% 
  rename("Monetary" = "MonetaryValue") %>% 
  select(Recency, Frequency, Monetary)

rownames(rfm_data) = rfm_table[,1]
```

## Plot the distribution of the recency, fequency and monetary variables

```{r plot-distribution-rfm}

par(mfrow = c(1, 3))  # Create a 1 row, 3 column layout



hist(rfm_data$Recency, ylab=NULL, xlab="Recency",probability = T, main = "RFM Recency Distribution", col="turquoise")


hist(rfm_data$Frequency, ylab=NULL,xlab="Frequency", main = "RFM Frequency Distribution", col="gold")

options(scipen = 999)
hist(rfm_data$Monetary, ylab=NULL,xlab="Monetary", main="RFM Monetary Distribtion", col="gray")


options(scipen = 0)
par(mfrow = c(1, 1))  

```

## Removing the skewness by performing log transformation on the variables
From the plot above we can see the data is skewed rightward, moreso for frequency and monetary values, the data is log transformed in order to reduce the impact of extreme values and make the data more symmetric and useful for analysis. The new plot shows the data normalised and centered.
```{r log-transformed-distribution, echo=FALSE}
# log transform the data to reduce skewness
rfm_log = rfm_data %>% 
  mutate(across(everything(), log))

rfm_log_plot = rfm_log %>% 
  pivot_longer(cols = everything(), names_to="Variable", values_to="Value")

# Histogram after log transformation
# ggplot(rfm_log_plot, aes(x = Value, fill = Variable)) +
#   geom_histogram(aes(y=after_stat(density)), binwidth = 1, position = "identity", alpha = 0.7) + 
#   labs(x = "Value", y = "Frequency", title = "Distribution of RFM after normalisation") +
#   theme_minimal()

par(mfrow = c(1, 3))  # Create a 1 row, 3 column layout

hist(rfm_log$Recency, ylab=NULL, xlab="Recency",probability = T, main = "RFM Recency Distribution", col="turquoise")

hist(rfm_log$Frequency, ylab=NULL,xlab="Frequency", main = "RFM Frequency Distribution", col="gold")

hist(rfm_log$Monetary, ylab=NULL,xlab="Monetary", main="RFM Monetary Distribtion", col="gray")

par(mfrow = c(1, 1))  # Create a 1 row, 3 column layout

```

## Standardizing the variables for equal variance and mean
Variables are standardized for several reasons in statistics. Standardization involves transforming variables to have a specific mean and standard deviation, often centered around 0 with a standard deviation of 1.
```{r normalised-data}
rfm_normalised = rfm_log %>% mutate(across(everything(), scale)) 
summary(rfm_normalised)
```

## Choosing number of Clusters using Elbow Method
This article on [r-bloggers](https://www.r-bloggers.com/2017/02/finding-optimal-number-of-clusters/) consicely explains the problem of finding the number of clusters.
The optimal cluster is chose
```{r clustering}
set.seed(1231)
k_values = 1:21

inertia = sapply(k_values, function(k){kmeans(rfm_normalised,centers = k )$tot.withinss})

plot(k_values, inertia,
     type="b", pch = 19, frame = FALSE, 
     xlim=c(1,20),
     xlab="Number of clusters K",
     ylab="Total within-clusters sum of squares", 
     main="Elbow Method", col="blue")

```

## Fit for 4 clusters

```{r four-clusters}
kmm.4 =  kmeans(rfm_normalised,4,nstart = 50,iter.max = 15)
cluster = kmm.4[["cluster"]]
rfm_4.cluster = cbind(rfm_data, cluster) %>% 
  group_by(cluster) %>% 
  summarise(Recency=mean(Recency), Frequency=mean(Frequency), Monetary=mean(Monetary))
(rfm_4.cluster)
```

## Fit for 5 clusters


While the 4 cluster segementation is simpler, 5 clusters provides more insight, better separating monetary values on the low and high ends.

1. Cluster 1: strong recency, moderate spenders
2. Cluster 2,4: weak recency and low spending
3. Cluster 3: moderate recency, moderate spending
4. Cluster 5: strong recency, frequency and high spending


```{r five-clusters}
kmm.5 =  kmeans(rfm_normalised,5,nstart = 50,iter.max = 15)
cluster = kmm.5[["cluster"]]
rfm_5.cluster = cbind(rfm_data, cluster) %>% 
  group_by(cluster) %>% 
  summarise(Recency=mean(Recency), Frequency=mean(Frequency), Monetary=mean(Monetary))
(rfm_5.cluster)
```


## Snake Plot
```{r snake-plot}
snake_plot_df = rfm_5.cluster %>% 
  pivot_longer(cols = c(Recency, Frequency, Monetary), 
               names_to="Metric",
               values_to= "Value")

ggplot(snake_plot_df, aes(x = Metric, 
                      y = Value, 
                      group = cluster, 
                      color=as.factor(cluster))) +
  geom_line() +
  labs(x = "Metric", y = "Value", color = "Cluster") +
  ggtitle("Line Plot with Multiple Lines") + 
  scale_y_continuous(labels = scales::comma)
```

## Finding the relative importance of each value

```{r relative-importance}
population_avg = rfm_data %>% summarise(across(everything(), mean)) 

relative_importance = rfm_5.cluster %>% 
  mutate(Recency = Recency/population_avg$Recency - 1,
         Frequency = Frequency/population_avg$Frequency - 1,
         Monetary = Monetary/population_avg$Monetary - 1)


ggplot(relative_importance %>% 
         pivot_longer(cols=2:4, names_to = "Variable", values_to="Value"), 
       aes(x = Variable, y = cluster, fill = Value)) +
  geom_tile() +
  scale_fill_gradient(low = "red", high = "green") +
  geom_text(aes(label = round(Value, 2)), color = "black") + 
  labs(title = "Heatmap", y = "Row", fill = "Value") + 
  theme(axis.title.x = element_blank())
```


